Comparing and combining network-guided biomarker discovery to discover genetic susceptibility mechanisms to breast cancer on the GENESIS study

### 2020-10-18 manuscript/plos/review_1.tex
updated latest version of the paper

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
% Network methods provide a comprehensive approach to uncovering the genetics of complex diseases and building hypotheses.
% They discover susceptibility genes by jointly consider the statistical association between genetic variation and a phenotype, measured in a genome-wide association study (GWAS), and the biological context of each gene, represented as a network.
Network approaches to disease use biological networks, which model functional relationships between the molecules in a cell, to generate hypotheses about the genetics of complex diseases.
Among them, several approaches jointly consider gene scores, representing the association between each gene and the disease, and the biological context of each gene, modeled by a network.
%These approaches can select low-scoring genes if they connect multiple high-scoring genes.
Here, we obtain gene scores from genome-wide association studies (GWAS) and study six network approaches on 
% Hence, a network method could select a gene with a high P-value of association if it connects multiple low P-value genes in the network. In this work, we studied six network methods which identify subnetworks with high scores of GWAS association with a phenotype. This allows them to give more compelling results than standard SNP- and gene-level GWAS analyses, recovering causal subnetworks tightly related to cancer susceptibility.
% We apply them to
GENESIS, a GWAS on French women with non-BRCA familial breast cancer% and no pathogenic variants in \emph{BRCA1} and \emph{BRCA2}
. We provide a critical comparison of these six methods, discussing the impact of their mathematical formulation and of their parameters. % on % . Additionally, we performed an in-depth benchmarking with respect to
% the size and predictive power of their solutions, as well as on their stability and runtimes.
% Importantly,
Using a biological network yields more compelling results than standard % SNP- and gene-level
GWAS analyses. Indeed, %recovering causal subnetworks tightly related to cancer susceptibility.
we find significant overlaps between our solutions and the genes identified in the largest GWAS on breast cancer susceptibility. % Since most of the methods produce reasonable solutions, we
We further
propose to combine these solutions into a consensus network, %containing the genes selected by at least two methods. This aggregation
which
brings further insights. The consensus network contains \emph{COPS5}, a gene related to multiple hallmarks of cancer, and 14 of its neighbors. The main drawback of network methods % with respect to conventional $\chi^2$-based GWAS
is that they are not robust to small perturbations in their inputs. %To compensate for this instability, we
We therefore propose a stable consensus solution, formed by the most consistently selected genes in multiple subsamples of the data. In GENESIS, it is composed of 68 genes, enriched in known breast cancer susceptibility genes (\emph{BLM}, \emph{CASP8}, \emph{CASP10}, \emph{DNAJC1}, \emph{FGFR2}, \emph{MRPS30}, and \emph{SLC4A7}, P\=/value~=~$3 \times 10^{-4}$) and occupying more central positions in the network than most genes. The network is organized around \emph{CUL3}, which is involved in the regulation of several genes linked to cancer progression. In conclusion, we showed how network methods help overcome the lack of statistical power of GWAS and improve their interpretation. Project-agnostic implementations of all methods are available at \url{https://github.com/hclimente/gwas-tools}. % to facilitate their application to other GWAS datasets.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}
In genome-wide association studies (GWAS), thousands of genomes are scanned to identify variants associated with a complex trait. Over the last 15 years, GWAS have advanced our understanding of the genetics of complex diseases, and in particular of hereditary cancers. Yet, they have led to an apparent paradox: the more we perform such studies, the more it seems that the entire genome is involved in every disease. The omnigenic model offers an appealing explanation: only a limited number of \emph{core} genes are directly involved in the disease; but gene functions are deeply interrelated, and so many other genes can alter the function of the core genes. These interrelations are often modeled as networks, and multiple algorithms have been proposed to use these networks to identify the subset of core genes involved in a specific trait. In this study, we apply and compare six such network methods on GENESIS, a GWAS dataset for familial breast cancer in the French population. Combining these approaches allows us to identify potentially novel breast cancer susceptibility genes, and provides a mechanistic explanation for their role in the development of the disease. We provide ready-to-use implementations of all the examined methods.
\linenumbers

@@

In human health, genome-wide association studies (GWAS) aim at quantifying how single-nucleotide polymorphisms (SNPs) predispose to complex diseases, like diabetes or some forms of cancer \cite{bush_chapter_2012}. To that end, in a typical GWAS thousands of unrelated samples are genotyped: the cases, suffering from the disease of interest, and the controls, taken from the general population. Then, a statistical test of association (e.g. based on logistic regression) is conducted between each individual SNP and the phenotype. Those SNPs with a P\=/value lower than a conservative Bonferroni threshold are candidates to further studies in an independent cohort. Once the risk SNPs have been discovered, they can be used for risk assessment, and to deepen our understanding of the disease.

GWAS have successfully identified thousands of variants underlying many common diseases \cite{buniello_nhgri-ebi_2019}. However, this experimental setting also presents intrinsic challenges. Some of them stem from the high dimensionality of the problem, as every GWAS to date studies more variants than samples are genotyped. This limits the statistical power of the experiment, as only variants with larger effects can be detected \cite{visscher_10_2017}. This is particularly problematic since the prevailing view is that most genetic architectures involve many variants with small effects \cite{visscher_10_2017}. Additionally, to avoid false positives, a conservative multiple test correction is applied, typically the previously mentioned Bonferroni correction. However, Bonferroni correction is overly conservative when the statistical tests are correlated, as it is the case in GWAS \cite{wang_statistical_2018}. Another open issue is the interpretation of the results, as the functional consequences of most common variants are unknown. On top of that, recent large-sampled studies suggest that numerous loci spread all along the genome contribute to a degree to any complex trait, in accordance with the infinitesimal model \cite{barton_infinitesimal_2017}. The recently proposed omnigenic model \cite{boyle_expanded_2017} offers an explanation: genes are strongly inter-related and influence each other's function, which allows alterations in most genes to impact the subset of ``core'' genes directly involved in the disease's mechanism. Hence, a comprehensive statistical framework which includes the structure of biological data might help alleviate the aforementioned issues.

For this reason, many authors turn to network biology to handle the complex interplay of biomolecules that lead to disease \cite{furlong_human_2013}. As its name suggests, network biology models biology as a network, where the biomolecules under study, often genes, are nodes, and selected functional relationships are edges that link them. These relationships come from evidence that the genes jointly contribute to a biological function; for instance, their expressions are correlated, or their products establish a protein-protein interaction. Under this view, complex diseases are not the consequence of a single altered gene, but of the interaction of multiple interdependent molecules \cite{barabasi_network_2011}. In fact, an examination of biological networks shows that disease genes have differential properties \cite{barabasi_network_2011,cai_broker_2010}: they tend to occupy central positions in the network (although not the most central ones); disease genes for the same pathology tend to be connected in modules; and often they are bottlenecks that interconect modules. Therefore, studying the neighborhood of disease-associated genes is effective at identifying new ones that are involved in the same biological functions \cite{huang_systematic_2018}.

Network-based discovery methods exploit the differential properties described above to identify disease genes on GWAS data \cite{azencott_network-guided_2016}. In essence, each gene is assigned a score of association with the disease, computed from the GWAS data, and biological relationships, given by a network built on prior knowledge. Then, the problem becomes finding a functionally-related set of highly-scoring genes. Multiple solutions have been proposed to this problem, often stemming from different mathematical frameworks and considerations of what the optimal solution looks like. For example, some methods restrict the problem to specific types of subnetworks. Such is the case of LEAN \cite{gwinner_network-based_2016}, which focuses on ``star'' subnetworks, i.e. instances were both a gene and its direct interactors are associated with the disease. Other algorithms, like dmGWAS \cite{jia_dmgwas:_2011} and heinz \cite{dittrich_identifying_2008}, do not impose such strong constraints, and search for subnetworks interconnecting genes with high association scores. However, they differ in their tolerance to the inclusion of low-scoring nodes, and the topology of the solution. Lastly, other methods also consider the topology of the network, favoring groups of nodes that are not only high-scoring, but also densely interconnected; such is the case of HotNet2 \cite{leiserson_pan-cancer_2015}, SConES \cite{azencott_efficient_2013}, and SigMod \cite{liu_sigmod:_2017}.

In this work, we studied the relevance of network-based approaches to genetics by applying these six network methods to GWAS data. They use different interpretations of the omnigenic model, and provide a representative view of the field. We worked on the GENESIS dataset \cite{sinilnikova_genesis:_2016}, a study on familial breast cancer conducted in the French population. After a classical GWAS approach, we used these network methods to identify additional breast cancer susceptibility genes. Lastly, we carried out a comparison of the solutions obtained by the different methods, and aggregated them to obtain consensus solutions of predisposition to familial breast cancer that addressed their shortcomings.

We conducted association analyses in the GENESIS dataset (Section~\ref{methods:data}) at both SNP and gene levels (Section~\ref{methods:conventional}). Two genomic regions had a P\=/value lower than the Bonferroni threshold on chromosomes 10 and 16 (\nameref{sfig:snp_gene_manhattan}A). The former overlaps with the gene \emph{FGFR2}; the latter with \emph{CASC16} and the protein-coding gene \emph{TOX3}. Variants in both \emph{FGFR2} and \emph{TOX3} have been repeatedly associated with breast cancer susceptibility in other case-control studies \cite{Michailidou2017}, \emph{BRCA1} and \emph{BRCA2} carrier studies \cite{Mulligan2011}, and in hereditary breast and ovarian cancer families negative for mutations in \emph{BRCA1} and \emph{BRCA2} \cite{rinella_genetic_2013}. In GENESIS only \emph{FGFR2} was significantly associated with breast cancer at the gene-level (\nameref{sfig:snp_gene_manhattan}B).

Closer examination reveals two other regions (3p24 and 8q24) having low, albeit not genome-wide significant, P\=/values. Both of them have been associated to breast cancer susceptibility in the past \cite{brisbin_meta-analysis_2011,search_newly_2009}. We applied an L1-penalized logistic regression using all GENESIS genotypes as input, and the phenotype (cancer/healthy) as outcome (Section~\ref{methods:classifier}). The algorithm selected 100 SNPs, both from all aforementioned regions and new ones (\nameref{sfig:snp_gene_manhattan}C). Yet, it is unclear why those SNPs were selected, as emphasized by the high P\=/value of some of them, which further complicates the biological interpretation. Moreover, and in opposition to what would be expected under the omnigenic model, the genes to which these SNPs map to (Section~\ref{methods:snp2gene}) are not interconnected in the protein-protein interaction network (PPIN, Section~\ref{methods:networks}). Moreover, the classification performance of the model is low (sensitivity = 55\%, specificity = 55\%, Section \ref{methods:benchmark}). Together, these issues motivate exploring network methods, which consider not only statistical association, but also the location of each gene in a PPIN to find susceptibility genes.

\begin{figure}[!ht]
  \centering
  \includegraphics[width=\textwidth]{./figures/figure_1.png}
  \caption{{\bf Overview of the solutions produced by the different network methods (Section~\ref{methods:methods}) on the GENESIS dataset. } As LEAN did not produce any significant solution (BH adjusted P\=/value~$<$~0.05), it was excluded. Unless indicated otherwise, results refer to genes, except for SConES GI which are at the SNP-level. \textbf{(A)}~Overlap between the genes selected by each of the methods, measured by Pearson correlation between indicator vectors (Sections \ref{methods:algorithm_comparison} and \ref{methods:snp2gene}). \textbf{(B)}~Distribution of VEGAS2 P-values of the genes in the PPIN not selected by any network method (12\,213), and of those selected by 1 (575), 2 (73), or 3 (20) methods. \textbf{(C)}~Solution networks produced by the different methods. \textbf{(D)}~Manhattan plots of SNPs/genes; in black, the method's solution. The Bonferroni threshold is indicated by a red line (2.54~\texttimes{}~10\textsuperscript{-7} for SNPs, 1.53~\texttimes{}~10\textsuperscript{-6} for genes).}
  \label{fig:solution_overview}
\end{figure}

These solutions differ in many aspects, making it hard to draw joint conclusions. For starters, the overlap between the genes featured in each solution is quite small (Fig~\ref{fig:solution_overview}A). However, the methods tend to agree on the genes with the strongest signal: genes selected by more methods tended to have lower P-value of association (Fig~\ref{fig:solution_overview}B).

Another prominent difference is the solution size: the largest solution, produced by HotNet2, contains 440 genes, while heinz's contains only 4 genes. While SConES GI did not recover any protein coding gene, working with SNP networks rather than gene networks allowed it to retriev four subnetworks in intergenic regions, and another one overlapping an RNA gene (\emph{RNU6-420P}).

The topologies of the five solutions differ as well (Fig~\ref{fig:solution_overview}C), as measured by the median centrality  and the number of connected components (Table~\ref{tab:gene_solutions}). Three methods yield more than one connected component: SConES, as described above, SigMod, and HotNet2. HotNet2 produced 135 subnetworks, 115 of which have fewer than five genes. The second largest subnetwork (13 nodes) contains the two breast cancer susceptibility genes \emph{CASP8} and \emph{BLM} (Section \ref{methods:bcac}).

\begin{table}[!ht]
  \begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
  \centering
  \caption{
  {\bf Summary statistics on the solutions of multiple network methods on the PPIN. The first row contains the summary statistics on the whole PPIN.}}
  \begin{tabular}{lrrrrrr}
  {\bf Network } & {\bf \# genes } & {\bf \# edges } & {\bf \# components } & {\bf \mean{Betweenness} } & {\bf \median{P}\textsubscript{gene} } & {\bf \# genes in consensus } \\
  \thickhline
  HINT HT           & 13\,619 & 142\,541  & 15  & 16\,706   & 0.46  & 93/93 \\
  \hline
  dmGWAS            & 194     & 450       & 1   & 49\,115   & 0.19  & 55/93 \\
  heinz             & 4       & 3         & 1   & 113\,633  & 0.001 & 4/93  \\
  HotNet2           & 440     & 374       & 130 & 7\,739    & 0.048 & 63/93 \\
  LEAN              & 0       & 0         & 0   & -         & -     & 0/93  \\
  SConES GI         & 0 (1)   & 0         & 0   & -         & -     & 0/93  \\
  SigMod            & 142     & 249       & 11  & 92\,603   & 0.008 & 84/93 \\
  \hline
  Consensus         & 93      & 186       & 21  & 50\,737   & 0.006 & 93/93 \\
  Stable consensus  & 68      & 49        & 32  & 94\,854   & 0.005 & 43/93 \\
  \end{tabular}
  \begin{flushleft} \textbf{\# genes}: number of genes selected out of those that are part of the PPIN; for SConES GI the total number of genes, including RNA genes, was added in parentheses. {\bf \# components }: number of connected components. \textbf{\mean{Betweenness}}: median betweenness of the selected genes in the PPIN. \textbf{\median{P}\textsubscript{gene}}: median VEGAS2 P\=/value of the selected genes. \textbf{\# genes in consensus}: number of genes in common between the method's solution, and the 93 genes in the consensus solution.
  \end{flushleft}
  \label{tab:gene_solutions}
  \end{adjustwidth}
\end{table}

Lastly, a pathway enrichment analysis (Section \ref{methods:pathway_enrichment}) also showed similarities and differences between the solutions of the different methods. It linked different parts of SigMod's solution to four processes (\nameref{stab:sigmod_pwy}): protein translation (including mitochondrial), mRNA splicing, protein misfolding, and keratinization (BH adjusted P\=/values~$<$~0.03). Interestingly, the dmGWAS solution (\nameref{stab:dmgwas_pwy}) was also related to protein misfolding (\emph{attenuation phase}, BH adjusted P\=/value~=~0.01). However, it additionally included submodules of proteins related to mitosis, DNA damage, and regulation of TP53 (BH adjusted P\=/values~$<$~0.05), which match previously known mechanisms of breast cancer susceptibility \cite{nielsen_hereditary_2016}. As with SigMod, the genes in HotNet2's solution (\nameref{stab:hotnet2_pwy}) were involved in mitochondrial translation (BH adjusted P\=/value~=~1.87 \texttimes{} 10\textsuperscript{-4}), but also in glycogen metabolism and transcription of nuclear receptors (BH adjusted P\=/value~$<$~0.04).

Despite their differences, there were additional common themes. All obtained solutions had lower association P\=/values than the whole PPIN (median VEGAS2 P\=/value $\ll 0.46$, Table~\ref{tab:gene_solutions}), despite containing genes with higher P\=/values as well (Fig~\ref{fig:solution_overview}D). This illustrates the trade-off between controlling for type I error and biological relevance. However, there were nuances between solutions in this regard: heinz strongly favored genes with lower P\=/values, while dmGWAS was less conservative (median VEGAS2 P\=/values 0.0012 and 0.19, respectively); SConES tended to select whole LD-blocks; and HotNet2 and SigMod were less likely to select low scoring genes.

Additionally, the solutions presented other desirable properties. First, four of them were enriched in known breast cancer susceptibility genes (dmGWAS, heinz, HotNet2, and SigMod, Fisher's exact test one-sided P\=/value$<$0.03). Second, the genes in three solutions displayed on average a significantly higher betweenness centrality than the rest of the genes (dmGWAS, HotNet2, and SigMod, Wilcoxon rank-sum test P\=/value $<$ 1.4 \texttimes{} 10\textsuperscript{-21}). This agrees with the notion that disease genes are more central than other non-essential genes \cite{pinero_uncovering_2016}, an observation that holds in breast cancer (one-tailed Wilcoxon rank-sum test P\=/value~=~2.64 \texttimes{} 10\textsuperscript{-5} when comparing the betweenness of known susceptibility genes versus the rest). Interestingly, SConES selected SNPs that are also more central than the average SNP (\nameref{stab:snp_solutions}), suggesting that causal SNPs are also more central than non-associated SNPs.

Despite their shared properties, the differences between the solutions of the different methods suggested that each of them captured different aspects of cancer susceptibility. Indeed, out of the 668 genes that were selected by at least one method, only 93 were selected by at least two, 20 by three, and none by four or more. Encouragingly, the more methods selected a gene, the higher its association score to the phenotype (Fig \ref{fig:solution_overview}B), a relationship that plateaus at 2. Hence, to leverage on their strengths and compensate their respective weaknesses, we built a consensus solution that captured the genes shared among at least two solutions (Section~\ref{methods:methods}). This solution (Fig~\ref{fig:consensus}) contained 93 genes and exhibited the aforementioned properties of the individual solutions: enrichment in breast cancer susceptibility genes and higher betweenness centrality than the rest of the genes.

\begin{figure}[!ht]
  \centering
  \includegraphics[width=.9\linewidth]{./figures/figure_3.pdf}
  \caption{ {\bf Consensus solution on GENESIS (Section~\ref{methods:methods}) }. \textbf{(A)} ~Manhattan plot of genes; in black, the ones in the consensus solution. The Bonferroni threshold is indicated by a red line (1.53~\texttimes{}~10\textsuperscript{-6} for genes). \textbf{(B)} Consensus network. Each gene is represented by a pie chart, which shows the methods that selected it. We enlarged the two most central genes (\emph{COPS5} and \emph{OFD1}) and those genes that are known breast cancer susceptibility genes and/or significantly associated with breast cancer susceptibility in the BCAC dataset (Section~\ref{methods:bcac}). \textbf{(C)} The nodes are in the same disposition as in panel B, but every gene name is indicated. We colored in pink the names of genes that are known breast cancer susceptibility genes and/or significantly associated with breast cancer susceptibility in the BCAC dataset.}
  \label{fig:consensus}
\end{figure}

A pathway enrichment analysis of the genes in the consensus solution also showed similar pathways as the individual solutions (\nameref{stab:consensus_pwy}). We found two involved mechanisms: \emph{mitochondrial translation} and \emph{attenuation phase}. The former is supported by genes like \emph{MRPS30} (VEGAS2 P\=/value~=~0.001), which encode a mitochondrial ribosomal protein and was also linked to breast cancer susceptibility \cite{quigley_5p12_2014}. Interestingly, increased mitochondrial translation has been found in cancer cells \cite{Yu2016Repositioning}, and its inhibition proposed as a therapeutic target. With regards to the attenuation phase of heat shock response, it involved three Hsp70 chaperones: HSPA1A, HSPA1B, and HSPA1L. The genes encoding these proteins are all near each other at 6p21, in the region known as HLA. In fact, out of the 22 SNPs that map to any of these three genes, 9 mapped to all of them, and 4 to two, making it hard to disentangle their effects. \emph{HSPA1A} was the most strongly associated gene (VEGAS2 P\=/value~=~8.37~\texttimes{}~10\textsuperscript{-4}).

Topologically, the consensus consisted of a connected component composed of 49 genes, and multiple smaller subnetworks (Fig~\ref{fig:consensus}B and C). Among the latter, 19 genes were in subnetworks containing a single gene or two connected nodes, implying that they did not have a consistently altered neighborhood, but were strongly associated themselves and hence picked by at least two methods. The large connected component contained genes that are highly central in the PPIN, a property which is weakly anti-correlated with the P\=/value of association to the disease (Pearson correlation coefficient~=~-0.26, \nameref{sfig:consensus_stats}). This suggested that these genes were selected because they were on the shortest path between another two highly associated genes. In view of this, we hypothesize that highly central genes might contribute to the heritability through alterations of their neighborhood, consistently with the omnigenic model of disease \cite{boyle_expanded_2017}. For instance, the most central node in the consensus solution is \emph{COPS5}, a component of the COP9 signalosome which regulates multiple signaling pathways. \emph{COPS5} is related to multiple hallmarks of cancer and is overexpressed in multiple tumors, including breast and ovarian cancer \cite{liu_jab1_cops5_2018}. Despite its lack of association in GENESIS or in studies conducted by the Breast Cancer Association Consortium (BCAC) \cite{Michailidou2017} (VEGAS2 P\=/value of 0.22 and 0.14 respectively), its neighbors in the consensus solution had consistently low P\=/values (median VEGAS2 P\=/value~=~0.006).

We compared the results obtained with different network methods to the European sample of BCAC, the largest GWAS to date on breast cancer (Section~\ref{methods:bcac}). Although BCAC case-control studies do not necessarily target cases with a familial history of breast cancer, this comparison is pertinent since we expect a shared genetic architecture at the gene level, at which most network methods operate. This shared genetic architecture, together with BCAC's scale (90 times more samples than GENESIS) provided a reasonable counterfactual of what we would expect if GENESIS had a larger sample size. We computed a gene association score on BCAC (Section~\ref{methods:bcac}). The solutions provided by the different  network methods overlapped significantly with BCAC hits (Fisher's exact test P\=/value \textless~0.019). The gene-based methods achieved comparable precision (2\%-25\%) and recall (1.3-12.1\%) at recovering BCAC-significant genes (\nameref{sfig:additional-benchmarks}A). Interestingly, while SConES GI, at the SNP-level, achieved a similar recall (8.6\%), it showed a much higher precision (47.3\%).

We compared the six network methods in a 5-fold subsampling setting (Section \ref{methods:benchmark}). This allowed us to measure five properties (Fig~\ref{fig:benchmark}): size of the solution; sensitivity and specificity of an L1-penalized logistic regression classifier on the selected SNPs; stability; and computational runtime. The solution size varies greatly between the different methods (Fig~\ref{fig:benchmark}A). Heinz produced the smallest solutions, with an average of 182 selected SNPs (Section~\ref{methods:snp2gene}). The largest solutions came from SConES GI (6\,256.6 SNPs), and dmGWAS (4\,255.0 SNPs). LEAN did not produce any solution in any of the subsamples. Using different combinations of parameters (Section \ref{methods:parameters}), we computed how good each of the methods was at recovering the results of a conventional GWAS on BCAC (Section \ref{methods:bcac}, Fig~\ref{fig:benchmark}B). SConES exhibits the largest area under the curve, since, when $\lambda = 0$ (network topology is disregarded), it is equivalent to a Bonferroni correction. Between the remaining network methods, they have similar areas under the curve, with heinz being the one with the largest area.

To determine whether the selected SNPs could be used for patient classification, we computed the performance of the classifier on \emph{test sets} (\nameref{sfig:additional-benchmarks}B). The different classifiers displayed similarly poor sensitivities and specificities, all in the 0.52 -- 0.56 range. Interestingly, the classifier trained on all the SNPs had a similar performance, despite being the only method aiming only at minimizing prediction error. Of course, although these performances were low, we did not expect to separate cases from controls well using exclusively genetic data \cite{deloscamposComplexTraitPredictionEra2018}.

\begin{figure}[!ht]
  \centering
  \includegraphics[width=.8\linewidth]{./figures/figure_4.pdf}
  \caption{{ \bf Comparison of network-based GWAS methods on GENESIS. } Each method was run 5 times on a random subset containing 80\% of the samples, and tested on the remaining samples (Section \ref{methods:benchmark}). As LEAN did not select any gene, it was excluded from all panels except \textbf{D}. \textbf{(A)}~Number of SNPs selected by each method and number of SNPs in the active set found by the classifier(Section~\ref{methods:classifier}). Points are the average over the 5 runs; the error bars represent the standard error of the mean. A grey diagonal line with slope 1 is added for comparison, indicating the upper bound of the active set (i.e. the number of SNPs in the solution). For reference, the active set of Lasso using all the SNPs included, on average, 154\,117.4 SNPs. \textbf{(B)} True positive rate and true negative rate, using significant SNPs (for SConES and $\chi^2$ + Bonferroni) and genes (for the remaining methods) in BCAC (Section~\ref{methods:bcac}) as true positives, of multiple parameter combinations for different methods (Section \ref{results:parameters}). \textbf{(C)}~Pairwise Pearson correlations of the solutions produced by different methods. A Pearson correlation of 1 means the two solutions are the same. A Pearson correlation of 0 means that there is no SNP in common between the two solutions. \textbf{(D)}~Runtime of the evaluated methods, by type of network used (PPIN or SNP). For gene-based methods, inverted triangles represent the runtime of the algorithm itself, and circles the total time, which includes the algorithm themselves and the additional 119\,980 seconds (1 day and 9.33 hours) that VEGAS2 took on average to compute the gene scores from SNP summary statistics.}
  \label{fig:benchmark}
  \end{figure}

Another desirable quality of a selection algorithm is the stability of the solution with respect to small changes in the input (Section~\ref{methods:algorithm_comparison}). Heinz was highly stable in our benchmark, while the other methods displayed similarly low stabilities (Fig~\ref{fig:benchmark}C).

In terms of computational runtime, the fastest method was heinz (Fig~\ref{fig:benchmark}D), which returned a solution in a few seconds. HotNet2 was the slowest (3 days and 14 hours on average). Including the time required to compute the gene scores, however, slowed down considerably gene-based methods; on this benchmark, that step took on average 1 day and 9.33 hours. Including this first step, it therefore took 5 days on average for HotNet2 to produce a result.

\subsection{Network topology and association scores matter, and might lead to ambiguous results}
\label{results:drawbacks}

As shown above, and despite their similarities, the network methods produced remarkably different solutions. This is due to the fact that each of them models the problem differently. Importantly, understanding which assumptions they made allowed to understand the results more in depth. For instance, the fact that LEAN did not return any gene implied that there was no gene such that both itself and its environment were on average strongly associated with the disease.

In the GENESIS dataset, heinz's solution was very conservative, providing a small solution with the lowest median P\=/value (Table~\ref{tab:gene_solutions}). By repeatedly selecting this compact solution, heinz was the most stable method (Fig~\ref{fig:benchmark}C). Its conservativeness stemmed from its preprocessing step, which modeled the gene P\=/values as a mixture model of a beta distribution and a uniform distribution, controlled by an FDR parameter. Due to the limited signal at the gene level in this dataset (\nameref{sfig:snp_gene_manhattan}B), only 36 of all the genes retained a positive score after that transformation. Yet, this small solution did not provide much insight into the susceptibility mechanisms to cancer. Importantly, it ignored genes that were associated to cancer in this dataset like \emph{FGFR2}.

On the other end of the spectrum, dmGWAS, HotNet2, and SigMod produced large solutions. dmGWAS' solution was the lowest scoring solution on average. This was due to the greedy framework it uses, which had a bias for larger solutions \cite{nikolayeva_network_2018}. It considered all nodes at distance 2 of the examined subnetwork, and accepted a weakly associated gene if it was linked to another, high scoring one. This was exacerbated when the results of successive greedy searches are aggregated, leading to a large, tightly connected cluster of unassociated genes (Fig~\ref{fig:issues}A). This relatively low signal-to-noise ratio combined with the large solution requires additional analyses to draw conclusions, such as enrichment analyses. In the same line, HotNet2's solution was even harder to interpret, being composed of 440 genes divided into 135 subnetworks. Lastly, SigMod missed some of the highest scoring, breast cancer susceptibility genes in the dataset, like \emph{FGFR2} and \emph{TOX3}.

>>>>

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Network approaches to disease use biological networks, which model functional relationships between the molecules in a cell, to generate hypotheses about the genetics of complex diseases. Several among them jointly consider gene scores, representing the association between each gene and the disease, and the biological context of each gene, modeled by a network. Here, we study six such network methods using gene scores from GENESIS, a genome-wide association study (GWAS) on French women with non-BRCA familial breast cancer. We provide a critical comparison of these six methods, discussing the impact of their mathematical formulation and parameters. Using a biological network yields more compelling results than standard GWAS analyses. Indeed, we find significant overlaps between our solutions and the genes identified in the largest GWAS on breast cancer susceptibility. We further propose to combine these solutions into a consensus network, which brings further insights. The consensus network contains \emph{COPS5}, a gene related to multiple hallmarks of cancer, and 14 of its neighbors. The main drawback of network methods is that they are not robust to small perturbations in their inputs. Therefore, we propose a stable consensus solution, formed by the most consistently selected genes in multiple subsamples of the data. In GENESIS, it is composed of 68 genes, enriched in known breast cancer susceptibility genes (\emph{BLM}, \emph{CASP8}, \emph{CASP10}, \emph{DNAJC1}, \emph{FGFR2}, \emph{MRPS30}, and \emph{SLC4A7}, P\=/value~=~$3 \times 10^{-4}$) and occupying more central positions in the network than most genes. The network is organized around \emph{CUL3}, which is involved in the regulation of several genes linked to cancer progression. In conclusion, we showed how network methods help overcome the lack of statistical power of GWAS and improve their interpretation. Project-agnostic implementations of all methods are available at \url{https://github.com/hclimente/gwas-tools}.

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}
Genome-wide association studies (GWAS) scan thousands of genomes to identify variants associated with a complex trait. Over the last 15 years, GWAS have advanced our understanding of the genetics of complex diseases, and in particular of hereditary cancers. However, they have led to an apparent paradox: the more we perform such studies, the more it seems that the entire genome is involved in every disease. The omnigenic model offers an appealing explanation: only a limited number of \emph{core} genes are directly involved in the disease, but gene functions are deeply interrelated, and so many other genes can alter the function of the core genes. These interrelations are often modeled as networks, and multiple algorithms have been proposed to use these networks to identify the subset of core genes involved in a specific trait. This study applies and compares six such network methods on GENESIS, a GWAS dataset for familial breast cancer in the French population. Combining these approaches allows us to identify potentially novel breast cancer susceptibility genes and provides a mechanistic explanation for their role in the development of the disease. We provide ready-to-use implementations of all the examined methods.
\linenumbers

@@

In human health, genome-wide association studies (GWAS) aim at quantifying how single-nucleotide polymorphisms (SNPs) predispose to complex diseases, like diabetes or some forms of cancer \cite{bush_chapter_2012}. To that end, in a typical GWAS, thousands of unrelated samples are genotyped: the cases, suffering from the disease of interest, and the controls, taken from the general population. Then, a statistical test of association (e.g., based on logistic regression) is conducted between each SNP and the phenotype. Those SNPs with a P\=/value lower than a conservative Bonferroni threshold are candidates to further studies in an independent cohort. Once the risk SNPs have been discovered, they can be used for risk assessment and deepening our understanding of the disease.

GWAS have successfully identified thousands of variants underlying many common diseases \cite{buniello_nhgri-ebi_2019}. However, this experimental setting also presents inherent challenges. Some of them stem from the high dimensionality of the problem, as every GWAS to date studies more variants than samples are genotyped. This limits the statistical power of the experiment, as it can only detect variants with larger effects \cite{visscher_10_2017}. This is particularly problematic since the prevailing view is that most genetic architectures involve many variants with small effects \cite{visscher_10_2017}. Additionally, to avoid false positives, most GWAS apply a conservative multiple test correction, typically the previously mentioned Bonferroni correction. However, Bonferroni correction is overly conservative when the statistical tests correlate, as happens in GWAS \cite{wang_statistical_2018}. Another open issue is the interpretation of the results, as the functional consequences of most common variants are unknown. On top of that, recent large-sampled studies suggest that numerous loci spread all along the genome contribute to a degree to any complex trait, in accordance with the infinitesimal model \cite{barton_infinitesimal_2017}. The recently proposed omnigenic model \cite{boyle_expanded_2017} offers an explanation: genes are strongly interrelated and influence each other's function, which allows alterations in most genes to impact the subset of ``core'' genes directly involved in the disease's mechanism. Hence, a comprehensive statistical framework that includes the structure of biological data might help alleviate the issues above.

For this reason, many authors turn to network biology to handle the complex interplay of biomolecules that lead to disease \cite{furlong_human_2013,leisersonNetworkAnalysisGWAS2013}. As its name suggests, network biology models biology as a network, where the biomolecules under study, often genes, are nodes, and selected functional relationships are edges that link them. These relationships come from evidence that the genes jointly contribute to a biological function; for instance, their expressions are correlated, or their products establish a protein-protein interaction. Under this view, complex diseases are not the consequence of a single altered gene, but of the interaction of multiple interdependent molecules \cite{barabasi_network_2011}. In fact, an examination of biological networks shows that disease genes have differential properties \cite{barabasi_network_2011,cai_broker_2010}: they tend to occupy central positions in the network (although not the most central ones); disease genes for the same pathology tend to cluster in modules; and often they are bottlenecks that interconnect modules.

Network-based discovery methods exploit the differential properties described above to identify disease genes on GWAS data \cite{azencott_network-guided_2016,huang_systematic_2018}. In essence, each gene receives a score of association with the disease, computed from the GWAS data, and a set of biological relationships, given by a network built on prior knowledge. Then, the problem becomes finding a functionally-related set of highly-scoring genes. Multiple solutions have been proposed to this problem, often stemming from different mathematical frameworks and considerations of what the optimal solution looks like. For example, some methods restrict the problem to specific types of subnetworks. Such is the case of LEAN \cite{gwinner_network-based_2016}, which focuses on ``star'' subnetworks, i.e., instances where both a gene and its direct interactors are associated with the disease. Other algorithms, like dmGWAS \cite{jia_dmgwas:_2011} and heinz \cite{dittrich_identifying_2008}, do not impose such strong constraints and search for subnetworks interconnecting genes with high association scores. However, they differ in their tolerance to the inclusion of low-scoring nodes and the topology of the solution. Lastly, other methods also consider the topology of the network, favoring groups of nodes that are not only high-scoring but also densely interconnected; such is the case of HotNet2 \cite{leiserson_pan-cancer_2015}, SConES \cite{azencott_efficient_2013}, and SigMod \cite{liu_sigmod:_2017}.

In this work, we studied the relevance of network-based approaches to genetics by applying these six network methods to GWAS data. They use different interpretations of the omnigenic model and provide a representative view of the field. We worked on the GENESIS dataset \cite{sinilnikova_genesis:_2016}, a study on familial breast cancer conducted in the French population. After a classical GWAS approach, we used these network methods to identify additional breast cancer susceptibility genes. Lastly, we compared the solutions obtained by the different methods and aggregated them to obtain consensus solutions of predisposition to familial breast cancer that addressed their shortcomings.
